\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-14 chapters/wcble.tex
Started processing comments MDV

The \ac{wcble} algorithm is presented in \cref{sec:wcble/app/wcble-alg}.
Its training phase is visualized in~\cref{fig:diagram-train}.
The initial training epochs $\{\mat{X}_n^{(1)}\}_{n=1}^N$ are set to $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N$.
At every iteration, classifier $\mathcal{C}$ is trained like in \ac{cble}:
\begin{equation}
  \theta^{(i)} =
  \mathrm{train}_\mathcal{C}(\{\mat{X}^{(i-1)}_n[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
\end{equation}
Next, latency $s_{l_n}^{(i)}$ is determined for every epoch $\mat{X}^{(i)}$ corresponding
to its class label $l_n$ using \cref{eq:prob-latency}.
Finally, the training epochs $\mat{X}^{(i+1)}$ for the next iteration are determined by aligning
each original training epoch to the latency $s_{l_n}^{(i)}$ corresponding to its respective class
label.
\begin{equation}
  \mat{X}^{(i+1)}_n = \mathrm{align}(\mat{X}_n^\mathrm{train}, s_{l_n}^{(i)}) \quad \forall n=1,\ldots,N
\end{equation}
Aligning is performed by shifting and zero-padding the signal to the right if
the latency is negative relative to the time window onset, and to the left if
positive, by the difference between the latency and the window onset.
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/figure1.pdf}
}{%
  Schematic representation of the \acs{wcble} training phase.
}{%
  Schematic representation of the \acf{wcble} training phase.
  (1) The first-stage spatiotemporal binary classifier is trained on a set of
  epochs.
  (2) It is then applied to time-shifted copies of the epochs to obtain scores
  and class probabilities over time.
  (3) The medians of these probability distributions are assumed as the class
  latencies.
  (4) The epochs are aligned to their corresponding class latencies by
  shifting in time such that all latencies fall at the same moment.
  (5) The spatiotemporal classifier is then retrained on the aligned epochs
  for the next iteration.
  (6) After the iterative process halts, the scores and latencies obtained
  from the last iteration are used to train the second-stage
  classifier.
}{fig:diagram-train}
The process halts after a fixed number of iterations or when the estimated set
of latencies has been encountered before, indicating it ended up in a loop.
In the end, the procedure should result in enhanced classifier parameters $\theta^*$,
closer to those when there would be no jitter between epochs.
Note that using the median approach for robust latency estimation results in a
smoother yet longer convergence process compared to the maximum/minimum approach.
We can then apply the classifier with enhanced parameters $\theta^*$ in a \ac{cble}
manner to unseen epochs as illustrated in \cref{fig:diagram-eval} to obtain a
vector of scores over time as in \cref{sec:wcble/methods/cble} and the estimated latencies as
in~\cref{sec:robust-latency}.
\fullpagefig{
  \includegraphics[width=\textwidth]{figures/wcble/figure2.pdf}

}{%
  Schematic representation of the \acs{wcble} test phase.
}{%
  Schematic representation of the \acf{wcble}. (1) The
	first-stage spatiotemporal binary classifier obtained from the training phase is
	applied to time-shifted copies of the epochs to obtain scores and class probabilities
	over time. (2) The medians of these probability distributions are assumed
	as the new class latencies. (3) The scores and class latencies are input to the
  trained second-stage classifier, which predicts the label of the epochs.
}{fig:diagram-eval}
\section{Results}
\label{sec:wcble/results}
Since it is hard to obtain ground truth measures of \ac{erp} latencies, we
first evaluate our approach with synthetic data.
In \cref{sec:covert-align}, the proposed algorithm will be applied to real
\ac{eeg} data.

@@

100 target epochs containing the evoked potential and noise and 100
non-target epochs containing only noise were simulated.
The target epochs were jittered by setting the latency offset $l$
of each target epoch to a random latency drawn from a normal distribution with 0 mean and standard
deviations of respectively $\sigma=0.1,0.2,0.3$s.
The noise was scaled at 32 different noise levels, with the \ac{snr} of a
target epoch ranging from 0 to -31dB, with
\begin{equation}
  \text{SNR} = 10\log_{10}\frac{\text{var}s(t,0)}{\text{var}\left(\text{noise}\right)}
\end{equation}

A sample of the simulated evoked data is displayed in \cref{fig:wcble/sim-sine}.
\fullpagefig{%
  \sffamily
  \sansmath
  \inputpgf{figures/wcble}{simulated-sine.pgf}
}{%
  Simulated target and non-target evoked data.
}{%
  Simulated target (bottom half of epochs) and non-target (top half of epochs) evoked data
  and their average (red: target, yellow: non-target, shaded area: standard
  deviation).
  Targets contain a sine wave pulse jittered by $\sigma$. 32 \ac{eeg} channels with pink
  spatiotemporal noise were simulated using a forward head model, and channel C3 is
  plotted here.
}{fig:wcble/sim-sine}

@@

As the \ac{snr} decreases, both \ac{cble} and \ac{wcble} can accurately
estimate evoked potential latencies up to a given threshold. After this
threshold, correlation between estimated and ground-truth latencies will start
decreasing.
However, we notice that this threshold occurs later for \ac{wcble} than for
\ac{cble} for all evaluated jitter levels, indicating that \ac{wcble} is more
robust in the presence of pink spatiotemporal noise.
At around -28dB, \ac{wcble} does not significantly outperform \ac{cble}
anymore, and both methods get overtaken by the strong presence of noise.

>>>>

The \ac{wcble} algorithm is presented in \cref{sec:wcble/app/wcble-alg}.
Its training phase is visualized in~\cref{fig:diagram-train}.
The initial training epochs $\{\mat{X}_n^{(1)}\}_{n=1}^N$ are set to $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N$.
At every iteration, classifier $\mathcal{C}$ is trained like in \ac{cble}:
\begin{equation}
  \theta^{(i)} =
  \mathrm{train}_\mathcal{C}(\{\mat{X}^{(i-1)}_n[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
\end{equation}
Next, latency $s_{l_n}^{(i)}$ is determined for every epoch $\mat{X}^{(i)}$ corresponding
to its class label $l_n$ using \cref{eq:prob-latency}.
Finally, the training epochs $\mat{X}^{(i+1)}$ for the next iteration are determined by aligning
each original training epoch to the latency $s_{l_n}^{(i)}$ corresponding to its respective class
label.
\begin{equation}
  \mat{X}^{(i+1)}_n = \mathrm{align}(\mat{X}_n^\mathrm{train}, s_{l_n}^{(i)}) \quad \forall n=1,\ldots,N
\end{equation}
Aligning is performed by shifting and zero-padding the signal to the right if
the latency is negative relative to the time window onset, and to the left if
positive, by the difference between the latency and the window onset.
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/figure1.pdf}
}{%
  \caption[Schematic representation of the \acs{wcble} training phase.]{%
    Schematic representation of the \acf{wcble} training phase.
    (1) The first-stage spatiotemporal binary classifier is trained on a set of
    epochs.
    (2) It is then applied to time-shifted copies of the epochs to obtain scores
    and class probabilities over time.
    (3) The medians of these probability distributions are assumed as the class
    latencies.
    (4) The epochs are aligned to their corresponding class latencies by
    shifting in time such that all latencies fall at the same moment.
    (5) The spatiotemporal classifier is then retrained on the aligned epochs
    for the next iteration.
    (6) After the iterative process halts, the scores and latencies obtained
    from the last iteration are used to train the second-stage
    classifier.
  }
  \label{fig:diagram-train}
}
The process halts after a fixed number of iterations or when the estimated set
of latencies has been encountered before, indicating it ended up in a loop.
In the end, the procedure should result in enhanced classifier parameters $\theta^*$,
closer to those when there would be no jitter between epochs.
Note that using the median approach for robust latency estimation results in a
smoother yet longer convergence process compared to the maximum/minimum approach.
We can then apply the classifier with enhanced parameters $\theta^*$ in a \ac{cble}
manner to unseen epochs as illustrated in \cref{fig:diagram-eval} to obtain a
vector of scores over time as in \cref{sec:wcble/methods/cble} and the estimated latencies as
in~\cref{sec:robust-latency}.
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/figure2.pdf}
}{%
  \caption[Schematic representation of the \acs{wcble} test phase.]{%
    Schematic representation of the \acf{wcble}. (1) The
	  first-stage spatiotemporal binary classifier obtained from the training phase is
	  applied to time-shifted copies of the epochs to obtain scores and class probabilities
	  over time. (2) The medians of these probability distributions are assumed
	  as the new class latencies. (3) The scores and class latencies are input to the
    trained second-stage classifier, which predicts the label of the epochs.
  }
  \label{fig:diagram-eval}
}
\section{Results}
\label{sec:wcble/results}
Since it is hard to obtain ground truth measures of \ac{erp} latencies, we
first evaluate our approach with synthetic data.
In \cref{sec:covert-align}, the proposed algorithm will be applied to real
\ac{eeg} data.

@@

% ERROR BETWEEN HERE ==========================================================
100 target epochs containing the evoked potential and noise and 100
non-target epochs containing only noise were simulated.
The target epochs were jittered by setting the latency offset $l$
of each target epoch to a random latency drawn from a normal distribution with 0 mean and standard
deviations of respectively $\sigma=0.1,0.2,0.3$s.
The noise was scaled at 32 different noise levels, with the \ac{snr} of a
target epoch ranging from 0 to -31dB, with
\begin{equation}
  \text{SNR} = 10\log_{10}\frac{\text{var}s(t,0)}{\text{var}\left(\text{noise}\right)}
\end{equation}

A sample of the simulated evoked data is displayed in \cref{fig:wcble/sim-sine}.
\fullpagefig{%
  \inputpgf{figures/wcble}{simulated-sine.pgf}
}{%
  \caption[Simulated target and non-target evoked data.]{%
    Simulated target (bottom half of epochs) and non-target (top half of epochs) evoked data
    and their average (red: target, yellow: non-target, shaded area: standard
    deviation).
    Targets contain a sine wave pulse jittered by $\sigma$. 32 \ac{eeg} channels with pink
    spatiotemporal noise were simulated using a forward head model, and channel C3 is
    plotted here.
  }
  \label{fig:wcble/sim-sine}
}

@@

% AND HERE ====================================================================
As the \ac{snr} decreases, both \ac{cble} and \ac{wcble} can accurately
estimate evoked potential latencies up to a given threshold. After this
threshold, correlation between estimated and ground-truth latencies will start
decreasing.
However, we notice that this threshold occurs later for \ac{wcble} than for
\ac{cble} for all evaluated jitter levels, indicating that \ac{wcble} is more
robust in the presence of pink spatiotemporal noise.
At around -28dB, \ac{wcble} does not significantly outperform \ac{cble}
anymore, and both methods get overtaken by the strong presence of noise.

### 2024-11-14 cover_kul.tex
Started processing comments MDV

\newgeometry{margin=1.5cm}
\begin{titlepage}

    \sffamily
    \begin{minipage}[t]{.3\textwidth}
    \vspace{0pt}
    KU Leuven \\
    Biomedical Sciences Group \\
    Faculty of Medical Science \\
    Department of Neurosciences

    \end{minipage}\hfill%
    \begin{minipage}[t]{.3\textwidth}
      \vspace{0pt}
      \includegraphics[width=\textwidth]{figures/kul_docschool.png}
      \bigskip

@@

    \vfill
    \begin{minipage}[b]{.5\textwidth}
      Jury: \\
      Supervisor KU Leuven: \emph{prof. Marc Van Hulle} \\
      Supervisor University of Lille: \emph{prof. François Cabestaing} \\
      Chair examining committee: \emph{prof. John Creemers} \\
      Chair public defence: \emph{prof. Koen Poesen} \\
      Jury members: \\
      \emph{prof. Maarten De Vos} \\
      \emph{prof. Adalberto Simeone} \\
      \emph{prof. Andrea K\"ubler} \\
      \emph{prof. Fabien Lotte} \\
      \emph{prof. Reinhold Scherer}
    \end{minipage}\hfill%
    \begin{minipage}[b]{.31\textwidth}\raggedright
        Dissertation presented in partial fulfillment of the requirements for
        the double degree of \emph{Doctor of Biomedical Sciences} (KU Leuven)
        and \emph{Doctor of Control Science and Signal Processing} (University
        of Lille)
    \end{minipage}
    \bigskip

    \centering\today

\end{titlepage}
\restoregeometry

>>>>

\begin{titlepage}
    \sffamily
    \begin{minipage}[t]{.5\textwidth}
    \vspace{0pt}
    KU Leuven \\
    Biomedical Sciences Group \\
    Doctoral School of Biomedical Sciences \\
    Faculty of Medical Science \\
    Department of Neurosciences
    \bigskip

    University of Lille \\
    Doctoral School MADIS \\
    UMR CRIStAL \\

    \end{minipage}\hfill%
    \begin{minipage}[t]{.3\textwidth}
      \vspace{0pt}
      \includegraphics[width=\textwidth]{figures/kul_docschool.png}
      \smallskip

@@

    \vfill
    \begin{minipage}[b]{.5\textwidth}
      \raggedright
      \small
      Jury: \\
      Co-supervisor KU Leuven: \emph{prof. Marc Van Hulle} \\
      Co-supervisor Univ. Lille: \emph{prof. François Cabestaing} \\
      Chair examining committee: \emph{prof. John Creemers} \\
      Chair public defense: \emph{prof. Koen Poesen} \\
      \emph{prof. Andrea K\"ubler} \\
      \emph{prof.\ Fabien Lotte} \\
      \emph{prof.\ Reinhold Scherer} \\
      \emph{prof.\ Maarten De Vos} \\
      \emph{prof.\ Adalberto Simeone}
    \end{minipage}\hfill%
    \begin{minipage}[b]{.31\textwidth}
        \raggedright
        \small
        Dissertation presented in partial fulfillment of the requirements for
        the double degree of \emph{Doctor of Biomedical Sciences} (KU Leuven)
        and \emph{Doctor of Control Science and Signal Processing} (University
        of Lille)
    \end{minipage}
    \bigskip

    \centering
    Defended December 16, 2024

\end{titlepage}
